### 2014-11-06 conclusions.tex
More corrections.

We think that our particular approach is applicable in distributed systems as we
have seen. Explore functional programming and coordination?

>>>>

We have presented a way of adding coordination to a declarative language without
change the nature of the language or introducing non-declarative constructs. We
took advantage of the fact that LM uses linear logic, which allows us to derive
coordination facts to delete facts in order to perform actions in the underlying
runtime system.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This is specially important since
programs can be proven to be correct even in the presence of coordination facts.
In order to prove our thesis, we first give a brief overview of the
We think that our particular approach is applicable in distributed systems as we
have seen. Explore functional programming and coordination?

### 2014-11-06 intro.tex
More updates.
More corrections.

We argue, however, that declarative programming leaves little to no programmer control
over how execution is scheduled or how data is laid out. This introduces some
performance issues because even if the runtime system is able to reasonably
parallelize the program using a general algorithm, there is a lack of specific
information about the program that a compiler does not easily understand. In
turn, such information would make execution better in terms of run time time,
memory usage and scalability.

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14}, with coordination facts that give programmer
control over scheduling and data placement. LM is a linear logic programming
language designed for programs that operate on graphs.  The use of linear
logic~\cite{girard-87} allows support for structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to communicate with other nodes.
Both computation and communication happen through derivation of logical rules.

Given that LM uses mutable state, the order in which nodes are executed may be
crucial in terms of performance. We extended the language with coordination
facts that can be used as a regular fact in the program logic to improve program
execution.  The first kind of coordination facts are called \emph{sensing facts}
and are used to sense information about the system the program is running on.
This allows the programmer to write logical rules that depend on the current
state of the program and the system. The second kind of coordination facts are
\emph{action facts} that when deleted in rules are used to apply a scheduling
operation during execution. This allows the programmer to prioritize node
computation or place nodes in different threads.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This is specially important since
programs can proven to be correct even in the presence of coordination facts.
In order to prove our thesis, we first give a brief overview of the
base language, including a small example. Next, in
Section~\ref{sec:coordination} we introduce the available coordination facts,
followed by Section~\ref{sec:implementation}, where we describe the
implementation of the runtime required to implement the desired coordination
mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
programmer control to declarative programs in order to make them perform better.

>>>>

The problem with declarative programming is that it leaves little to no programmer control
over how execution is scheduled or how data is laid out, making it hard to achieve
necessary efficiency. This introduces some
performance issues because even if the runtime system is able to reasonably
parallelize the program using a general algorithm, there is a lack of specific
information about the program that a compiler does not easily understand. In
turn, such information would make execution better in terms of run time time,
memory usage and scalability.

In this paper, we extend Linear Meld (LM), a declarative language by
\cite{cruz-iclp14,cruz-ppdp14} with coordination facts that give programmer
control over scheduling and data placement. LM is a linear logic programming
language designed for programs that operate on graphs.  The use of linear
logic~\cite{girard-87} allows support for structured manipulation of mutable
state. In LM, computation is divided so that each node of the graph computes
independently and is allowed to communicate with other nodes.
Both computation and communication happen through derivation of logical rules.

Given that LM uses mutable state, the order in which nodes are executed may be
crucial in terms of performance. We extended the language with coordination
facts that can be used as a regular fact in the program logic to improve program
execution.  The first kind of coordination facts are called \emph{sensing facts}
and are used to sense information about the system the program is running on,
e.g., scheduling and node placement on threads. The second kind of coordination facts are
\emph{action facts} that when deleted in rules are used to apply a scheduling
operation during execution. Coordination facts allow the programmer to write
logical rules that depend on the current state of the program and then prioritize node
computation or place nodes in different threads.

To the best of our knowledge, this is the first time that a declarative language allows 
control over execution without a significant change in the language itself and
without using non-declarative operations. This is specially important since
programs can be proven to be correct even in the presence of coordination facts.
In order to prove our thesis, we first give a brief overview of the
base language, including a small example. Next, in
Section~\ref{sec:coordination} we introduce the available coordination facts,
followed by Section~\ref{sec:implementation}, where we describe the
implementation of the runtime required to implement the desired coordination
mechanisms. Afterwards, in Section~\ref{sec:applications} we present several applications where we add
programmer control to declarative programs in order to make them perform better.

### 2014-11-06 language.tex
Small changes.
More corrections.

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate and its associated tuple of values, representing the
arguments. Since LM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated with new and
deleted facts. When no more rules are applicable, the program terminates.

LM has been designed for writing programs that operate on graphs. LM partitions
the database by using the first argument of each fact. The first argument is
typed as a \emph{node} and represents a node in the graph. For instance, the
fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
is stored in node \texttt{@2}. LM restricts
the body of every rule to only refer to the same node so that nodes can derive
rules independently. Although the body is restricted, the head of the
rule may refer to any node as long as the node is refered somewhere in the
body. This allows communication between nodes during rule derivation, since a
node must send a fact to another node.
Rule restrictions in turn make LM implicitly parallel because nodes are able to
compute independently as long as they have new facts to be applied in rules.
This makes LM non-deterministic since nodes can be picked to run in any order,
affecting which rules are applied and which facts are deleted or derived.

@@

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, is a persistent
predicate that describes the
relationship between the nodes of the graph, where the third argument represents
the weight of the edge
(the \texttt{route} modifier informs the compiler about the structure of the
program graph). The predicates \texttt{shortest} and \texttt{relax} are specified
as linear facts and thus are deleted when deriving new facts.
The idea of the algorithm is to compute the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a \texttt{shortest}
fact that is improved with new \texttt{relax} facts.
Lines 5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance (infinity)
for all nodes; and \texttt{relax(@1, 0, [@1])} starts the algorithm with the
initial distance from \texttt{@1} to \texttt{@1}.

@@

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      {B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ [B])}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o path(A, D2, P2).
\end{Verbatim}
  \caption{Single Source Shortest Path program code.}
  \label{code:shortest_path_program}
\end{figure}
\normalsize

The first rule of the program (lines 11-15) replaces the current path in
\texttt{shortest} with a shorter one in \texttt{relax}. The rule deletes both
\texttt{relax} and \texttt{shortest} facts and derives a new \texttt{shortest}
fact. In lines 14-15, we have a \emph{comprehension} where the program iterates
over the edges of node \texttt{A} and derives a new \texttt{relax} fact at
\texttt{B} with the new shorter distance plus the weight of the edge. For
instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

@@

The second rule of the program (lines 17-19) deals with the case where the new
distance in \texttt{relax} is not better than the current one, so it is thrown
away and the current distance is kept.

>>>>

Linear Meld (LM) is a forward-chaining linear logic programming
language~\cite{cruz-iclp14} with roots on Datalog~\cite{Ramakrishnan93asurvey}.
Programs consist of a set of \emph{rules} and a \emph{database of facts}.
Rules such as \texttt{a(X), b(Y) -o c(X, Y)} can be read as follows: if fact \texttt{a(X)}
and fact \texttt{b(Y)} exist in the database then \texttt{c(X, Y)} is added to
the database. The expression \texttt{a(X), b(Y)} is called the \emph{body} of
the rule and \texttt{c(X, Y)} is the \emph{head} of the rule.
A fact is a predicate, e.g., \texttt{a}, \texttt{b} or \texttt{c}, and its
associated tuple of values, e.g., the values of \texttt{X} and \texttt{Y}, representing the
arguments. Since LM uses linear logic as its foundation, we distinguish between
\emph{linear} and \emph{persistent facts}. Linear facts are deleted during the
process of deriving rule, while persistent facts are not.
Program execution starts by adding the \emph{axioms} (the initial facts) of the program to the database.
Next, the rules are recursively applied and the database is updated by adding
new facts or deleting facts used during rule derivation.
When no more rules are applicable, the program terminates.

\newcommand{\scare}[1]{``#1''}

LM has been designed for writing programs that operate on graphs. LM partitions
the database by using the first argument of each fact. The first argument is
typed as a \emph{node} and represents a node in the graph. For instance, the
fact \texttt{f(@1, 2)} is stored in node \texttt{@1}, while fact \texttt{p(@2)}
is stored in node \texttt{@2}. LM restricts
the body of every rule to only refer to the same node so that nodes can derive
rules independently. Although the body is restricted, the head of the
rule may refer to any node as long as the node is refered somewhere in the
body. This allows \scare{communication} between nodes during rule derivation, since a
node may \scare{send} a fact to another node.
Rule restrictions in turn make LM implicitly parallel because nodes are able to
compute independently. This makes LM non-deterministic since nodes can be picked
to run in any order, affecting which rules are applied and which facts are
deleted or derived.

@@

The SSSP program in Fig.~\ref{code:shortest_path_program} starts (lines 1-3)
with the declaration of the predicates. Predicates specify the kinds of facts
used in the program. The first predicate, \texttt{edge}, is a persistent
predicate that describes the relationship between the nodes of the graph,
where the third argument represents the weight of the edge (the
\texttt{route} modifier informs the compiler that the \texttt{edge} predicate
determines the structure of the graph).
The predicates \texttt{shortest} and \texttt{relax} are
specified as linear facts and thus are deleted when deriving new facts.  The
algorithm computes the shortest distance from node
\texttt{@1} to all other nodes in the graph. Every node has a
\texttt{shortest} fact that is improved with new \texttt{relax} facts.  Lines
5-9 declare the axioms of the program: \texttt{edge} facts describe the
graph; \texttt{shortest(A, +00, [])} is the initial shortest distance
(infinity) for all nodes; and \texttt{relax(@1, 0, [@1])} starts the
algorithm with the initial distance from \texttt{@1} to \texttt{@1}.

@@

shortest(A, D1, P1), D1 > D2, relax(A, D2, P2)
   -o shortest(A, D2, P2),
      {B, W | !edge(A, B, W) |
         relax(B, D2 + W, P2 ++ [B])}.

shortest(A, D1, P1), D1 <= D2, relax(A, D2, P2)
   -o shortest(A, D1, P1).
\end{Verbatim}
  \caption{Single Source Shortest Path program code.}
  \label{code:shortest_path_program}
\end{figure}
\normalsize

The first rule of the program (lines 11-14) reads as following: if the current
\texttt{shortest} path \texttt{P1} with distance \texttt{D1} is larger
than a new path \texttt{relax} with distance \texttt{D2}, then replace the
current shortest path with \texttt{D2}, delete the new \texttt{relax} path and
propagate new paths to the neighbors (lines 14-15) using a \emph{comprehension}.
The comprehension iterates over the edges of node \texttt{A} and derives a new
\texttt{relax} fact for each node \texttt{B} with the distance \texttt{D2 + W},
where \texttt{W} is the weight of the edge. For
instance, in Fig.~\ref{fig:shortest_path_program}~(a) we apply rule 1 in node
\texttt{@1} where two new \texttt{relax} facts are derived at node \texttt{@2}
and \texttt{@3}. Fig.~\ref{fig:shortest_path_program}~(b) is the result after
applying the same rule but at node \texttt{2}.

@@

The second rule of the program (lines 16-17) is read as following: if the
current shortest path \texttt{D1} is shorter than the new path \texttt{D2} then
delete the new \texttt{relax} fact and keep the old shortest path.

### 2014-11-06 nqueens.tex
More corrections.

The LM solution considers the chess board as a graph where nodes communicate
with each other to build valid configurations. First, we start with the empty
state in the first row of the chess board. Then, each square adds its own
position to the state and sends the state down, to the next row. Once a square
receives new configurations, it attempts to add its position to the
configurations. If valid, that configuration is then sent, recursively, to the
next row, until all rows are traversed. At the end of the program, the nodes at
the bottom row will have all the valid configurations.

Since computation goes from the top row to the bottom row, not all placements of
nodes on threads may be equally performant. This is specially true because the
bottom rows tend to perform most work. The best placement is then to split the
board vertically, so that, in optimal conditions, each thread gets the same
number of columns.  Our experiments show that, on shared memory, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by improving load balancing dynamically. But what if we want run the
program on a distributed system, where moving nodes between workers can be
expensive? We measured the performance of the program without node stealing in
order to assess how important data placement would be for such systems.
We used the default partition of the program as it is, which results in the board being
split by rows and another configuration using the coordination action
\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
the coordinates of the node in the chessboard.

>>>>

The LM solution considers the chess board as a graph where nodes exchange valid
configurations between each other. First, we start with the empty
state in the first row of the chess board. Then, each square adds its own
position to the state and sends the state down, to the next row. Once a square
receives new configurations, it attempts to add its position to the
configurations. If valid, that configuration is then sent, recursively, to the
next row, until all rows are traversed. At the end of the program, the nodes at
the bottom row will have all the valid configurations.

Since computation goes from the top row to the bottom row, not all placements of
nodes on threads may be equally performant. This is specially true because the
bottom rows tend to perform the most work. The best placement is then to split the
board vertically, so that, in optimal conditions, each thread gets the same
number of columns.  Our experiments show that, on shared memory, it does not
matter much if we start with a bad placement since node stealing overcomes those
issues by improving load balancing dynamically. But what if we want run the
program on a distributed system, where moving nodes between workers can be
expensive? We measured the performance of the program without node stealing in
order to assess how important data placement would be for such systems.
We used the default partition of the program as it is, which results in the board being
split by rows and another configuration using the coordination action
\texttt{set-cpu(A, vertical(X, Y))} that forces nodes to be moved to another
thread using a vertical partitioning. Note that \texttt{X} and \texttt{Y} are
the coordinates of the node in the chessboard.

### 2014-11-06 related.tex
More corrections.

There has been an increasing interest in declarative and data-centric
languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
popular data-centric programming model that is optimized for large clusters. The
scheduling and data sharing model is very simple: in the \emph{map phase}, data
is transformed at each node and the result reduced to a final result in the
\emph{reduce phase}. In order to facilitate the writing of programs over large
datasets, SQL-like languages such as
PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
PigLatin automatically compiles declarative code into physical plans
that are executed over Hadoop but provides very little control to the
programmer.

A declarative approach that is regaining popularity is
Datalog~\cite{Ullman:1990:PDK:533142}, also a forward-chaining logic programming
language. Traditionally used in deductive databases, it is now being increasingly used in different fields
such as distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
LM itself is inspired in P2~\cite{Loo-condie-garofalakis-p2} since it also models
programs as a graph structure. Another programming system that uses the graph
abstraction is Dryad~\cite{Isard:2007:DDD:1272996.1273005}, which combines computational vertices
with communication channels (edges) to form a data-flow graph. The program is
then scheduled to run on multiple computers/cores and data is partitioned
automatically during runtime.

@@

Another important coordination language is Delirium~\cite{Delirium}. Unlike
Linda which is embbeded into another language, Delirium actually embeds
operators written in other languages inside the Delirium language.
The advantages of Delirium are improved abstraction
and easier debugging because sequential operators are isolated from the
coordination language.

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. LM also differs from those languages since it
raises the abstraction level from the processing units to nodes of a graph.
Furthermore, the LM language is both a coordination language and a computation
language and there is no distinction between the two components.

\iffalse
GraphLab~\cite{GraphLab2010} is a C++ framework for developing parallel machine
learning algorithms. GraphLab allows nodes to
have read/write access to different scopes through different concurrent access
models in order to balance performance and data consistency. While some programs
only need to access the local node's data, others may need to update edge
information. Each consistency model will provide different guarantees that are
better adapted to some algorithms. GraphLab provides different schedulers
that dictate the order in which node's are computed, which is a rudimentary form
of coordination. Later in this paper, we will show how certain GraphLab's
schedulers can be easily implemented in LM through the use of coordination
facts.
\fi

>>>>

There has been an increasing interest in declarative and data-centric
languages. MapReduce~\cite{Dean:2008:MSD:1327452.1327492}, for instance, is a
popular data-centric programming model that is optimized for large clusters. The
scheduling and data sharing model is very simple: in the \emph{map phase}, data
is transformed at each node and the result reduced to a final result in the
\emph{reduce phase}. In order to facilitate the writing of programs over large
datasets, SQL-like languages such as
PigLatin~\cite{Olston:2008:PLN:1376616.1376726} have been developed.
PigLatin builds on top of MapReduce and allows the programmer to write complex
data-flow graphs, increasing the abstraction of the MapReduce paradigm.
An alternative to PigLatin/MapReduce is
Dryad~\cite{Isard:2007:DDD:1272996.1273005}
that allows programmers to design arbitrary computation patterns using the DAG
abstraction. It combines computational vertices with communication channels
(edges). Dryad programs are automatically scheduled to run on multiple
computers/cores and data is partitioned automatically during runtime.

@@

Another early coordination language is Delirium~\cite{Delirium}. Unlike
Linda which is embbeded into another language, Delirium actually embeds
operators written in other languages inside the Delirium language.
The advantages of Delirium are improved abstraction
and easier debugging because sequential operators are isolated from the
coordination language.

Linda and Delirium are limited in the sense that the programmer can only
coordinate the scheduling of processing units, while the placement of data is
left to the implementation. LM differs from those languages since it
is fully data-driven and coordination acts on data instead of processing units.
The abstraction is then raised by considering data and algorithmic aspects of
the program instead of focusing how processing units are used.
Furthermore, the LM language is both a coordination language and a computation
language and there is no distinction between the two components.

Forward-chaining logic programming was made famous by the Datalog
language~\cite{Ullman:1990:PDK:533142}.  Traditionally used in deductive
databases, it is now being increasingly used in different fields such as
distributed networking~\cite{Loo-condie-garofalakis-p2}, sensor
nets~\cite{Chu:2007:DID:1322263.1322281} and cloud computing~\cite{alvaro:boom}.
LM itself is inspired in Meld~\cite{ashley-rollman-iclp09}, a Datalog-like language for
programming distributed emsembles of modular robots. Meld introduced the idea of
sensing and action facts in order to sense and act on the outside world,
respectively. Coordination facts are inspired in the same principles but are
used for coordinating data and computation.

### 2014-11-06 sssp.tex
More corrections.

The Single Source Shortest Path (SSSP) program was already presented in
Section~\ref{sec:language}. We are going to update the code in
Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
are selected to run before others nodes. The coordinated version of the same
algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 15). Note that we
also use a global program directive to order priorities in ascending order (line
5).

@@

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      \{B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ [B]),
         \underline{set-priority(B, float(D1 + W))}\}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o shortest(A, D2, P2).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program_coord}
\end{figure}
\normalsize

>>>>

The Single Source Shortest Path (SSSP) program was already presented in
Section~\ref{sec:language}. We are going to update the code in
Fig.~\ref{code:shortest_path_program} so that nodes with the shortest distances
are selected to run before others nodes. The coordinated version of the same
algorithm is presented in Fig.~\ref{code:shortest_path_program_coord} and it
uses one coordination fact, namely \texttt{set-priority} (line 14). Note that we
also use a global program directive to order priorities in ascending order (line
5).

@@

shortest(A, D1, P1), D1 > D2, relax(A, D2, P2)
   -o shortest(A, D2, P2),
      \{B, W | !edge(A, B, W) |
         relax(B, D2 + W, P2 ++ [B]),
         \underline{set-priority(B, float(D2 + W))}\}.

shortest(A, D1, P1), D1 <= D2, relax(A, D2, P2)
   -o shortest(A, D1, P1).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program_coord}
\end{figure}
\normalsize
